The Natural History of Child Signals of Need in Utila, Honduras: An Exploratory Study

### 2024-10-18 Paper.qmd
Editing results for new figures. More exploratory analyses of anthropometrics

```{r setup, include=FALSE}
source("dictionaries.R", local = knitr::knit_global())
source("analysisforpaper.R", local = knitr::knit_global())

```

@@

|  |  |
|----|----|
| Variable | Survey question wording |
| Crying Frequency | How often does your child cry? |
| Sad Frequency | How often is your child sad? |
| Tantrum Frequency | How often does your child throw temper tantrums? (e.g., scream, throws things, or throws themselves on the floor) |
| Running Away Frequency | How often does your child leave the house without telling you when they get mad? |
| Parent/Child Conflict Frequency | How often do you have conflicts with your child? |
| Child Alloparenting Frequency | How often does your child take care of their other siblings? |

@@

|  |  |
|----|----|
| Variable | Survey question wording |
| Relative Need | How much time, effort, and resources does this child need compared to your other children? |
| Relative Investment | How much time, effort, and resources do you put into this child compared to your other children? |

@@

@@

<!-- [NOTE: Add categorical variables to the table; combine child and adult variables into 1 table] -->
```{r sumtab_children}
#| label: tbl-sumtab_children
#| tbl-cap: Summary statistics for child variables.

@@

::: {#fig-PCA}
```{r}
#| fig-width: 11
plot_pca
```

@@

@@

## Signaling behavior is negatively associated with age

As children mature, they are more capable of fulfilling their own needs, more able to ask others for what they need, and are less vulnerable to threats. We therefore expected signaling frequencies to decrease with age. For every signaling measure except sadness (which also has non-signaling functions), there was a statistically significant negative relationship between the age of the child and the frequency of signaling. See @stbl-main-models and @fig-effect_age.

Age had the largest effect on crying (`{r} stats$mcryH$ChildAge$str`), with the model predicting that by age 8 crying frequency drops to half of its value at age 5 (visual inspection). The effect of age on temper tantrums (`{r} stats$mtantrumH$ChildAge$str`) was smaller with a more linear slope.

::: {#fig-effect_age}
```{r}
signaling_data_plot_age
```

The effect of age on specific signal frequencies, summed signal frequency, the frequency of a child's most common signal, and the summed cost of signaling behavior, all per month. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Rug corresponds to actual values of child age with a small amount of jitter to reveal overlapped values. See @stbl-main-models for the corresponding regression table.
:::

## Signaling behavior is positively associated with conflict with primary caretakers

Costly signals of need are necessary when there are conflicts of interest. There were statistically significant positive relationships between the frequency of conflict between a child and their primary caretaker and all of our signaling measures (although frequent signaling might cause caretaker perceptions of conflict; see the Discussion section). See @stbl-main-models and @fig-effect_conflict.

::: {#fig-effect_conflict}
```{r}
signaling_data_plot_conflict
```

The effect of the frequency of conflict with one's primary caretaker on specific signal frequencies, summed signal frequency, the frequency of a child's most common signal, and signal cost, all per month. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Rug corresponds to actual values of the frequency of conflict between child and primary caretaker with a small amount of jitter to reveal overlapped values. See @stbl-main-models for the corresponding regression table.
:::

## Signaling behavior is negatively associated with caregiver education

\[delete this section if we remove conflict from the main model? or mention on significant when controlling for conflict\]

We did not predict relationships between child signaling and our demographic control variables, but we did find statistically significant negative relationships between years of caregiver education and all of our signaling measures except sadness (`{r} stats$msadH$EducationLevelYears$str`). See @stbl-main-models and @fig-effect_edu.

::: {#fig-effect_edu}
```{r}
signaling_data_plot_education
```

The effect of years of caregiver education on specific signal frequencies, summed signal frequency, the frequency of a child's most common signal, and signal cost, all per month. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Rug corresponds to actual values of primary caretaker education (years) with a small amount of jitter to reveal overlapped values. See @stbl-main-models for the corresponding regression table.
:::

## Signaling behavior and child alloparenting effort by sex

High alloparenting effort by children is associated with conflict with their parents (Syme and Hagen, xx). Across cultures, girls do more alloparenting than boys (xx, yy). We therefore investigated the association of child alloparenting effort and child signaling, with an interaction term for child sex. For each signaling measure, there was also a significant interaction between how often a child cared for their siblings and the sex of the child. The slope of these relationships were positive for girls and flat or slightly negative for boys. In other words, our models predict that as children provide more allocare within their families, the girls’ frequency of signaling increases with increases in their alloparenting effort, and boys’ frequency of signaling remains at the same rate or decreases slightly with increases in their alloparenting effort. See @stbl-main-models and @fig-effect_allo_by_sex.

::: {#fig-effect_allo_by_sex}
```{r}
signaling_data_plot_allo_by_sex
```

The effect of an alloparenting effort by sex interaction on specific signal frequencies, summed signal frequency, the frequency of a child's most common signal, and the summed cost of signaling behavior, all per month. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Rug corresponds to actual values of child alloparenting frequency with a small amount of jitter to reveal overlapped values. See @stbl-conflict-dep-var for the corresponding regression table.
:::

## Predictors of conflict between children and caretakers

Because caretaker perceptions of conflict predicted child signaling frequency, we investigated predictors of conflict. The frequency of conflict between children and caretakers is negatively associated with child age (`{r} stats$mconflict$ChildAge$str`) and the number of adults who provide childcare in a household (`{r} stats$mconflict$AdultsChildcare$str`) when controlling for the sex of the child, household demographics, the number of adults who provide and do not contribute to childcare, an alloparenting effort by sex interaction, and the summed frequency of alloparenting from all other children in the household. See @stbl-conflict-dep-var and @fig-effect_conflict_dep_var.

::: {#fig-effect_conflict_dep_var}
```{r}
signaling_data_plot_conflict_dep_var
```

The effect of A) child age and B) the number of adults in the household that contribute to childcare on the frequency of conflict between caretaker and child. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Rug corresponds to actual values of A) child age and B) the number of adults in the household that contribute to childcare with a small amount of jitter to reveal overlapped values. See @stbl-conflict-dep-var for the corresponding regression table.
:::

## Negative association between child alloparenting effort and number of older children

\[DELETE THIS SECTION OR MOVE TO SI???\]

Because child allparenting effort was associated with child signaling, we investigated predictors of child alloparenting effort. We expected that children’s alloparenting effort would be positively associated with their age, female sex, and the number of younger siblings, and negatively with the number of older siblings and the total alloparenting effort of others in the household. Initial linear modeling of child alloparenting effort revealed that a few households with exceptionally high levels of total child alloparenting effort produced models that predicted extremely unrealistic levels of child alloparenting effort. We therefore removed `{r}`r sum(d2$OtherChildAlloparentingFreqN > 121, na.rm = TRUE)` outliers (`{r} `r sum(d2$OtherChildAlloparentingFreqN == 360, na.rm = TRUE)`of which had a value of 360 times per month and`{r} `r sum(d2$OtherChildAlloparentingFreqN == 420, na.rm = TRUE)` of which had a value of 420 timers per month) despite expecting these values to not be outliers in the broader Utilian community.

::: {#fig-effect_alloparenting_dep_var_plot}
```{r}
alloparenting_dep_var_plot
```

The effect of the number of older children in the household on the frequency of a child's alloparenting effort. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Each dot is 1 child. Overlapping points are represented by the transparency of the point (lower transparency equals more values at a given coordinate). See @stbl-alloparenting-dep-var for the corresponding regression table.
:::

@@

|  |  |
|----|----|
| Variable | Survey question wording |
| IllnessAvoidance1 | If an illness is 'going around', my child will get it. |
| IllnessAvoidance2 | In general, my child is very susceptible to colds, flu, and other infectious diseases |
| IllnessAvoidance3 | My child’s immune system protects him/her from most illnesses that other kids get. |

>>>>

```{r setup, include=FALSE}
source("dictionaries.R", local = knitr::knit_global())
source("analysisforpaper.R", local = knitr::knit_global())
# source("regularize.R")
```

@@

|  |  |
|------------------------------------|------------------------------------|
| Variable | Survey question wording |
| Crying Frequency | How often does your child cry? |
| Sad Frequency | How often is your child sad? |
| Tantrum Frequency | How often does your child throw temper tantrums? (e.g., scream, throws things, or throws themselves on the floor) |
| Running Away Frequency | How often does your child leave the house without telling you when they get mad? |
| Parent/Child Conflict Frequency | How often do you have conflicts with your child? |
| Child Alloparenting Frequency | How often does your child take care of their other siblings? |

@@

|  |  |
|------------------------------------|------------------------------------|
| Variable | Survey question wording |
| Relative Need | How much time, effort, and resources does this child need compared to your other children? |
| Relative Investment | How much time, effort, and resources do you put into this child compared to your other children? |

@@

The overall structure of study variables was analyzed with principle components analysis (PCA) and Bayesian graphical models (BGM) and minimum spanning trees (MST).

@@

<!-- [NOTE: Add categorical variables to the table; combine child and adult variables into 1 table] -->

```{r sumtab_children}
#| label: tbl-sumtab_children
#| tbl-cap: Summary statistics for child variables.

@@

::: {#fig-PCA}
![](Figures/plot_pca.svg)

@@

We then estimated a BGM for all study variables, extracting the MST, which revealed that signaling variables clustered together, conflict was positively associated with signal cost, and child age negatively associated with crying frequency. See @fig-mst.

::: {#fig-mst}
![](Figures/plot_mst_uw.svg)

Minimum spanning tree of child signaling and variables thought to cause differences in signaling strategies. Blue edges = negative \[do we want more detail here? e.g., partial correlations or w/e the proper Bayesian way to word that is\] correlations between variables. Red edges = positive correlations between variables. Edge lengths represent \[check that this is right and if more specificity is needed\] distances between variables. Line weights represent(1 - correlation coefficient)
:::

@@

## Predictors of signaling frequency and conflict frequency

Using lasso regression, we explored all potential predictors of each child signal frequency, as well as of conflict frequency. See @fig-lasso.

::: {#fig-lasso}
![](Figures/plot_lasso.svg)

Lasso regression models of each child signaling frequency and conflict frequency. Variables whose coefficient values were shrunk to 0 are not displayed. Models fit uins using quasi-Poisson regression.
:::

The effects of particularly large risk or protective factors for child signaling are shown in @fig-signal-effects.

::: {#fig-signal-effects}
![](Figures/signal_effects_plot.svg)

Selected predictors of child signaling frequencies.
:::

The effects of particularly large risk or protective factors for conflict frequency are shown in @fig-conflict-effects.

::: {#fig-conflict-effects}
![](Figures/conflict_effects_plot.svg)

Selected predictors of child-parent conflict frequencies.
:::

@@

|  |  |
|------------------------------------|------------------------------------|
| Variable | Survey question wording |
| IllnessAvoidance1 | If an illness is 'going around', my child will get it. |
| IllnessAvoidance2 | In general, my child is very susceptible to colds, flu, and other infectious diseases |
| IllnessAvoidance3 | My child’s immune system protects him/her from most illnesses that other kids get. |
